Mining Scenario/Transaction Pairs in HW/SW Co-verification Using Symbolic Execution

### 2017-04-06 paper.tex
composition figure

%===============================================================================
\begin{abstract}
%===============================================================================
%
Time-to-market pressure requires early development of firmware and other
drivers for circuitry well before RTL is finished.  A key risk is divergence
of the hardware RTL from the behavior expected by the software.  Formal
methods can deliver a proof that the RTL and the firmware match, but are
difficult to scale up.  We propose to identify partitions of the firmware and
the RTL that follow transaction boundaries by means of path-based symbolic
simulation with custom path-pruning.  We have implemented this approach in
our tool \verifox.  We have experimentally compared \verifox with a
conventional approach, which constructs a single monolithic SAT/SMT
instance, and observe speedup of an order of magnitude over conventional 
approach for proving safety properties as well as detecting firmware/hardware bugs.
%
\end{abstract}

@@

In the post-RTL phase, when the RTL design is available, hardware/software
co-verification becomes more complex.  Unlike the TLM, which only captures
limited design functionality, the RTL implements cycle-accurate behavior
of the final HW, and addresses concerns related to power, area and
timing.  The effectiveness of testing 
is thus severely limited in post-RTL co-verification. Verification is mandatory to ensure that the driver or firmware (FW) 
is executed correctly together with the RTL.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %                    Challenges in Co-verification 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The verification of a FW given in C/C++ together with an RTL given in a hardware
description language (HDL) such as Verilog/VHDL is a challenging task due to a
few reasons -- 1) the timing and semantic mismatch between the synchronous
clock-driven HW model and asynchronous event-driven SW model. For example, the
FW running on a processor could be much faster than the HW
model it interacts with, 2) lack of support for property specification in 
HW/FW co-designs that can precisely capture the design intent of the 
communication between HW and FW, 3) large
number of interleavings between HW and FW threads, and 4) lack 
of automated formal tools that support co-designs implemented in C/C++ and 
Verilog/VHDL.   In this paper, we address some of these challenges to enable
seamless adoption of formal techniques in hardware/software co-verification.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %                     Handling Semantic Gap  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Generating a unified FW-HW model for co-verification is a 
well-known phenomenon to handle the semantic mismatch between  
HW~\footnote{Note that whenever we refer to ``hardware'' 
from this point onwards, we mean an RTL implementation 
and not a TLM.}  and FW models.  Most of the existing techniques 
in~\cite{fmsd02,emsoft07,fase10} for constructing a unified FW-HW model are not 
useful in practise for three reasons -- 1) use of abstract HW models 
instead of RTL implementation, 2) the HW-FW model
construction is agnostic to the underlying reasoning engine, and 3) 
it does not leverage the property specification over HW-FW boundary, hence 
ineffective for practical co-verification.  We will call the unified FW-HW 
model as \emph{co-verification model}.

@@

A desirable characteristic of a co-verification model is the ability to exploit 
the expressiveness of the underlying reasoning engines.  To this end, \hwcbmcv
constructs either a bit-level co-verification model that is checked with a SAT solver 
or a word-level model that is checked with an SMT solver.  Whereas, a 
co-verification model in \verifox is constructed 
by first translating the RTL description into a software representation in C,
called \emph{software netlist}, which is then followed by sequentializing the FW 
model and the software netlist model.  This results in a single
threaded sequential co-verification model which is represented in C. A path-wise 
symbolic execution is then performed over the co-verification model to
automatically mine relevant interactions between the FW and HW models.

@@

\caption{Proposed HW/SW Co-verification Flow

@@

\textbf{Step 3: Co-verification Model}
%
An unified co-verification model is constructed through appropriate composition of 
FW and its interacting HW threads.  This is done by instrumenting the 
software netlist model of HW RTL inside the FW thread at every invocation 
of the $next\_timeframe()$ primitive in interface functions.  Recall that a
call to $next\_timeframe()$ is used to advance the HW clock by one cycle.  A call
to the software netlist emulates the effect of one clock period in the RTL 
model. That is, the state of state holding elements (registers) and combinational 
elements (wires) are uniform in software netlist and RTL model after every invocation of 
$next_timeframe()$.  Furthermore, a software
netlist is a sequential model of HW RTL which is represented in C.  Thus, the
composition leads to a single-threaded sequential co-verification model in C.  
%
We now formally define a our co-verification model, denoted by 
$\mathcal{M}_{seq}$.
%
\[ 
\begin{array}[t]{@{}lll}
  \mathcal{M}_{seq} & {:}{:}{=} & \mathit{Procedure} \\
\mathit{Procedure} & {:}{:}{=} & \mathit{Statement} \mid \mathit{Procedure} \\
\mathit{Statement} & {:}{:}{=} & CStatement \mid \mathit{moduleName(Var_1,\dots,Var_n)} \\
\mathit{CStatement} & {:}{:}{=} & x {:}{=} exp \mid \mathit{ITE}(b, s_1,s_2) \mid \mathit{s_1;s_2} \mid loop\{s\} \\ 
\end{array}
\]
%
Consider sets of expressions $Exp$ and Boolean expressions $BExp$ 
over variables $Var$ of $\mathcal{M}_{seq}$ such that $(exp \in Exp)$ and 
$(b \in BExp)$. A $\mathit{CStatement}$ is an assignment,
conditional, sequential concatenation or a loop.
$\mathit{moduleName(Var_1,\dots,Var_n)}$ invokes the top level 
procedure of $\mathcal{SN}$, which is also the top level 
module of the HW RTL, since the translation from RTL to software preserves
the module hierarchy.  The interface variables $\{Var\_1,\dots,Var\_n\}$ 
(input, output and inout ports) are declared inside the 
FW to communicate with $\mathcal{SN}$.  Figure~\ref{compose} shows the 
composition of $\mathcal{SN}$ and FW thread for the UART IP.
%
%\begin{figure}[t]
%\scalebox{.50}{\import{figures/}{compose.pspdftex}}
%  \caption{Composition of FW and HW threads}
%\label{fig:interleaving2}
%\end{figure}
%

\textbf{Step 4: Property Driven Slicing of Co-verification Model}
A property-driven slicing is performed on the unified model, $\mathcal{M}_{seq}$.  
This step is purely syntactic, meaning that we perform a backward dependency 
analysis starting from the property which only preserve those 
program fragments that are relevant to a given property. The sliced program 
is then passed to the symbolic execution engine for path-based exploration. \\
%
%=======================================================================================
%\subsection{\verifox: Path-wise Symbolic Execution Tool} 
%=======================================================================================
\input{verifox}
%
\subsection{Approach 2: Monolithic Co-verification Technique}
%
\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.6]{figures/traditional_flow.pdf}
\caption{Monolithic HW/SW Co-verification Flow}
\label{fig:conventional}
\end{center}
\end{figure}
%
Figure~\ref{fig:conventional} illustrates the monolithic approach to HW/SW
co-veri\-fi\-cation, as implemented in \hwcbmcv.  
%\hwcbmcv is the only open-source
%formal co-verification tool that supports Verilog RTL (System Verilog 2009
%standard) and C/C++ firmware.  
In contrast to path-based approach,
\hwcbmcv maintains two separate flows for HW and FW.  The top flow in
Figure~\ref{fig:conventional} uses synthesis to obtain either a bit-level or
a word-level netlist from Verilog RTL.  A bit-level netlist is represented 
in And Inverter Graph (AIG) format.  Whereas, a word-level netlist is represented 
in format that resembles SMT-LIB standards.  The bottom flow illustrates the
translation of C program into static single assignment (SSA) form. 
Synthesizing netlist from RTL is a standard technique which is widely 
used in hardware model checkers~\cite{}. Similarly, translating a C program 
into a SSA representation is a standard technique which is widely used in 
software verification tools, such as CBMC~\cite{}, 2LS~{}. 
These two flows meet only at the solver.  Thus, \hwcbmcv generates a
unified monolithic co-verification model from AIG and SSA which is 
represented in Conjunctive Normal Form (CNF).  This model is then used to perform 
bounded model checking~\cite{} using the underlying SAT/SMT solvers.  
\hwcbmcv provides specific handshake primitives such
as $next\_timeframe()$ and $set\_inputs()$ to model FW-HW communications.

\rmcmt{
To this end, \hwcbmcv constructs a co-verification model through 
in tandem symbolic execution of the HW and FW models.  A bit-level 
netlist representation in And Inverter Graph (AIG) format or word-level netlist in
format similar to SMT-LIB standards is synthesized from a HW Register Transfer 
Level (RTL) implementation. Similarly, the SW, given in C, is automatically 
translated into a Static Single Assignment (SSA) form.  A monolithic verification 
condition is generated from these two models which is then solved using a 
}
%
\Omit{
The hardware and firmware run independently of each other. 
The communication between them takes place through these function calls. 
}
%===============================================================================
\section{Property Specification for Co-verification}\label{property}
%===============================================================================
Lack of support for property specification in a HW/SW co-design is one of the
stumbling blocks for the application of formal techniques in co-verification.  
In this paper, we express a co-design property in C 
with the support of special primitive, called {\em next\_timeframe()}, 
that can express temporal correlation between HW and FW events. 
In this paper, we classify co-design properties into two 
classes -- {\em transaction-level} property and {\em component-level} property.
%-------------------------------------------------------------------------------
\subsection{Transaction-level Property}
%-------------------------------------------------------------------------------
A {\em Transaction-level} property ($\psi_{t}$) formally captures the 
design intent of the interaction between a $\mathcal{ST}$ and $\mathcal{HT}$, 
that is, it is defined at the transaction boundaries.  
Recall that a $\mathcal{ST}$ is a service in response to some external environment 
or HW interrupts (for example acknowledgement or exceptions). Similarly, 
$\mathcal{HT}$ is a finite-length service in response to some SW events or 
other HW events.

>>>>

%===============================================================================
\begin{abstract}
%===============================================================================
%
Time-to-market pressure requires early development of firmware and other
drivers for circuitry well before RTL is finished.  A key risk is divergence
of the hardware RTL from the behavior expected by the software.  Formal
methods can deliver a proof that the RTL and the firmware match, but are
difficult to scale up.  We propose to identify partitions of the firmware and
the RTL that follow transaction boundaries by means of path-based symbolic
simulation with custom path-pruning.  We have implemented this approach in
our tool \verifox.  We have experimentally compared \verifox with a
conventional approach, which constructs a single monolithic SAT/SMT
instance, and observe speedups of an order of magnitude over conventional 
approach for proving safety properties as well as detecting firmware/hardware bugs.
%
\end{abstract}

@@

In the post-RTL phase, when the RTL design is available, hardware/software
co-verification becomes more complex.  Unlike the TLM, which only captures
limited design functionality, the RTL implements cycle-accurate behavior
of the final HW, and addresses concerns related to power, area and
timing.  Verification is mandatory to assert that the driver or firmware (FW) 
is executed correctly together with the RTL.  The effectiveness of testing 
is thus severely limited in post-RTL co-verification.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %                    Challenges in Co-verification 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The verification of FW given in C/C++ together with the RTL given in a hardware
description language (HDL) such as Verilog/VHDL is a challenging task due to 
few reasons -- 1) the timing and semantic mismatch between the synchronous
clock-driven HW model and asynchronous event-driven SW model -- for example,
FW running on a processor could be much faster than the HW
model it interacts with, 2) lack of support for property specification in 
HW/FW co-designs that can precisely capture the design intent of the 
communication between HW and FW, 3) Large
number of interleavings between HW and FW threads, and 4) lack 
of automated formal tools that support co-designs implemented in C/C++ and 
Verilog/VHDL.   In this paper, we address some of these challenges to enable
seamless adoption of formal techniques in hardware/software co-verification.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %                     Handling Semantic Gap  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Generating a unified FW-HW model for co-verification is a 
well-known phenomenon to handle the semantic mismatch between the 
HW~\footnote{Note that whenever we refer to ``hardware'' 
from this point onwards, we mean an RTL implementation 
and not a TLM.}  and FW models.  Most of the existing techniques 
in~\cite{fmsd02,emsoft07,fase10} for constructing a unified FW-HW model are not 
useful in practise for three reasons -- 1) use of abstract HW models 
instead of register-transfer level implementation, 2) the HW-FW model
construction is agnostic to the underlying reasoning engine, and 3) 
does not leverage the property specification over HW-FW boundary, hence 
ineffective for practical co-verification.  We will call the unified FW-HW 
model as \emph{co-verification model}.

@@

A desirable characteristic of a co-verification model is the ability to exploit 
the expressivity of the underlying reasoning engines.  To this end, \hwcbmcv
constructs either a bit-level co-verification model that are checked with a SAT solvers 
or a word-level model that are checked with SMT solvers.  Whereas, a 
co-verification model in \verifox is constructed 
by first translating the RTL description into a software representation in C,
called \emph{software netlist}, which is then followed by sequentializing the FW 
model and the software netlist model.  This results in a single
threaded sequential co-verification model which is represented in C. A path-wise 
symbolic execution is then performed over the co-verification model to
automatically mine relevant interaction behaviors between the FW and HW models.

@@

\caption{Path-based HW/SW Co-verification Flow in \verifox

@@

\textbf{Step 3: Co-verification Model}
%
An unified co-verification model is constructed through appropriate composition of 
FW and its interacting HW threads.  This is done by instrumenting the 
software netlist model of HW RTL inside the FW thread at every invocation 
of the $next\_timeframe()$ primitive in interface functions.  Recall that a
call to $next\_timeframe()$ is used to advance the HW clock by one cycle.  A call
to the software netlist emulates the effect of one clock period in the RTL 
model. That is, the state of state holding elements (registers) and combinational 
elements (wires) are uniform in software netlist and RTL model after every invocation of 
$next_timeframe()$.  Furthermore, a software
netlist is a sequential model of HW RTL which is represented in C.  Thus, the
composition leads to a single-threaded sequential co-verification model in C.  
%
We now formally define a our co-verification model, denoted by 
$\mathcal{M}_{seq}$.
%
\[ 
\begin{array}[t]{@{}lll}
  \mathcal{M}_{seq} & {:}{:}{=} & \mathit{Procedure} \\
\mathit{Procedure} & {:}{:}{=} & \mathit{Statement} \mid \mathit{Procedure} \\
\mathit{Statement} & {:}{:}{=} & CStatement \mid \mathit{moduleName(Var_1,\dots,Var_n)} \\
\mathit{CStatement} & {:}{:}{=} & x {:}{=} exp \mid \mathit{ITE}(b, s_1,s_2) \mid \mathit{s_1;s_2} \mid loop\{s\} \\ 
\end{array}
\]
%
Consider sets of expressions $Exp$ and Boolean expressions $BExp$ 
over variables $Var$ of $\mathcal{M}_{seq}$ such that $(exp \in Exp)$ and 
$(b \in BExp)$. A $\mathit{CStatement}$ is an assignment,
conditional, sequential concatenation or a loop.
$\mathit{moduleName(Var_1,\dots,Var_n)}$ invokes the top level 
procedure of $\mathcal{SN}$, which is also the top level 
module of the HW RTL, since the translation from RTL to software preserves
the module hierarchy.  The interface variables $\{Var\_1,\dots,Var\_n\}$ 
(input, output and inout ports) are declared inside the 
FW to communicate with $\mathcal{SN}$.  Figure~\ref{compose} shows the 
composition of $\mathcal{SN}$ and FW thread for the UART IP.
%
\begin{figure}[t]
\scalebox{.50}{\import{figures/}{compose.pspdftex}}
  \caption{Composition of FW and HW threads}
\label{compose}
\end{figure}

\textbf{Step 4: Property Driven Slicing of Co-verification Model}
A property-driven slicing is performed on the unified model, $\mathcal{M}_{seq}$.  
This step is purely syntactic, meaning that we perform a backward dependency 
analysis starting from the property which only preserve those 
program fragments that are relevant to a given property. The sliced program 
is then passed to the symbolic execution engine for path-based exploration. \\
%
%=======================================================================================
%\subsection{\verifox: Path-wise Symbolic Execution Tool} 
%=======================================================================================
\input{verifox}
%
\subsection{Approach 2: Monolithic Co-verification Technique}
%
\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.6]{figures/traditional_flow.pdf}
\caption{Monolithic HW/SW Co-verification Flow in \hwcbmcv}
\label{fig:conventional}
\end{center}
\end{figure}
%
\textbf{Step 1: Co-verification Model}
%
Figure~\ref{fig:conventional} illustrates the monolithic approach to HW/SW
co-veri\-fi\-cation, as implemented in \hwcbmcv.  
%\hwcbmcv is the only open-source
%formal co-verification tool that supports Verilog RTL (System Verilog 2009
%standard) and C/C++ firmware.  
In contrast to path-based approach,
\hwcbmcv maintains two separate flows for HW and FW.  The top flow in
Figure~\ref{fig:conventional} uses synthesis to obtain either a bit-level or
a word-level netlist from Verilog RTL.  A bit-level netlist is represented 
in And Inverter Graph (AIG) format.  Whereas, a word-level netlist is represented 
in format that resembles SMT-LIB standards.  The bottom flow illustrates the
translation of C program into static single assignment (SSA) form. 
Synthesizing netlist from RTL is a standard technique which is widely 
used in hardware model checkers~\cite{DBLP:journals/tse/Clarke76}.
Similarly, translating a C program into a SSA representation is a 
standard technique which is widely used in 
software verification tools, such as CBMC~\cite{cbmc.tacas:2004}. 
\hwcbmcv generates a monolithic co-verification model from AIG and SSA through 
in-tandem symbolic execution of HW and FW components.

\textbf{Step 2: Bounded Model Checking}
%
The co-verification model is checked with bounded model checking (BMC).
Given an unwind depth depth~$k$ and a co-design property $P$, BMC operates
by unwinding the co-verification model~$M$ up to depth~$k$ starting from
initial state $x_0$, represented by an initial state predicate~$I$.
This results in the following formula which is then checked for
satisfiability using an efficient SAT or SMT procedure.
%
\[ I(x_0) \wedge \bigwedge_{i=0}^{k-1} M(x_i, x_{i+1}) \wedge \bigvee_{i=0}^{k} 
\neg P(x_i) \]
%
Thus, BMC exploits the finiteness of $M$ and creates k copies of the
$M$ for each unrolling.

%\hwcbmcv provides specific handshake primitives such
%as $next\_timeframe()$ and $set\_inputs()$ to model FW-HW communications.

\textbf{Step 3: Scenario/Transaction Pair} In \hwcbmcv, the symbolic execution 
of HW and FW components meet only at the solver phase.  Thus, a complex
monolithic formula is passed on to the solver for verification purpose.  This
limits the opportunity for extracting relevant scenario/transaction pairs
through symbolic execution. Instead, the technique relies on the underlying
SAT/SMT solver to effeciently reason about the co-verification model. 
%
\Omit{
The hardware and firmware run independently of each other. 
The communication between them takes place through these function calls. 
}
%===============================================================================
\section{Property Specification for Co-verification}\label{property}
%===============================================================================
Lack of support for property specification in a HW/SW co-design is one of the
stumbling blocks for the application of formal techniques in co-verification.  
In this paper, we express a co-design property in C 
with the support of special primitive, called {\em next\_timeframe()}, 
that can express temporal correlation between HW and FW events. 
In this paper, we classify co-design properties into two 
classes -- {\em transaction-level} property and {\em component-level} property.
%-------------------------------------------------------------------------------
\subsection{Transaction-level Property}
%-------------------------------------------------------------------------------
A {\em Transaction-level} property ($\psi_{t}$) formally captures the 
design intent of the interaction between a $\mathcal{ST}$ and $\mathcal{HT}$, 
that is, it is defined at the transaction boundaries.  
Recall that a $\mathcal{ST}$ is a service in response to some external environment 
or HW interrupts (for example acknowledgement or exceptions). Similarly, 
$\mathcal{HT}$ is a finite-length service in response to some SW events or 
other HW events.
